Clustering time series based on probability distributions across temporal granularities

### 2021-11-17 main.Rmd
all references fixed

\noindent This work is motivated by a need to cluster a large collection of residential smart meter data, so that households can be grouped into similar energy usage patterns. These can be considered to be univariate time series of continuous values which are available at fine temporal scales. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and sporadic missing values. Using probability distributions is a natural way to analyze this types of data because they are robust to uneven length, missing data, or noise. This paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across a selection of cyclic granularities. Cyclic temporal granularities [@Gupta2021-hakear], which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can measure repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across cyclic granularities. The benefits of this approach are as follows.

@@

The primary application of this work is data from the Smart Grid, Smart City (SGSC) project (2010–2014) available through the [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). Half-hourly measurements of usage for more than 13,000 household electricity smart meters is provided from from October 2011 to March 2014. <!--Larger data sets include greater uncertainty about customer behavior due to growing variety of customers.--> Households vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some households use a dryer, while others dry their clothes on a line. Their weekly usage profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Daily energy usage varies depending on whether customers stay home or work away from home. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior complex. The challenge is to be able to cluster consumers into these type of expected patterns, and other unexpected patterns, using only their energy usage history [@Ushakova2020-rl]. <!--To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner.--> There is a growing need to have methods that can examine the energy usage heterogeneity observed in smart meter data and what are some of the most common power consumption patterns.

@@

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of households showing similar periodic behavior by considering $wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Cluster characterization is an important final stage of a cluster analysis. The primary purpose is to compare the homogeneity within a cluster to the heterogeneity of clusters. This can be done numerically, by tabulating cluster means and standard deviations [@dasu2005grouping], and visually using methods for graphics multivariate data. @Cook2007-qe provides visual examples using both tours [@asimov] and parallel coordinate plots [@wegman1990hyperdimensional]. Dimension reduction techniques like principal component analysis [@joliffe], multidimensional scaling (MDS) [@borg2005modern], t-distributed stochastic neighbor embedding (t-SNE) [@maarten-tsne] and linear discrimant analysis (LDA) [@lda] are also useful. <!-- A Parallel Coordinates Plot features parallel axes for each variable and each axis is linked by lines. Changing the axes may reveal patterns or relationships between variables for categorical variables. However, for categories with cyclic temporal granularities, preserving the underlying ordering of time is more desirable. Displaying cluster statistics is useful when we have larger problems and it is difficult to read the parallel coordinate plots due to congestion. All of MDS, PCA and t-SNE use a distance or dissimilarity matrix to construct a reduced-dimension space representation, but their goals are diverse. Multidimensional scaling [@borg2005modern) seeks to maintain the distances between pairs of data points, with an emphasis on pairings of distant points in the original space. The t-SNE embedding will compress data points that are close in high-dimensional space. Tour is a collection of interpolated linear projections of multivariate data into lower-dimensional space. The cluster characterization approach varies depending on the distance metric used. Parallel coordinate plots, scatter plot matrices, MDS or PCA are potentially useful ways to characterize clusters using wpd-based distances. For JS-based distances, plotting cluster statistics is beneficial for characterization and variable importance could be displayed through parallel coordinate plots. -->

@@

<!-- 1. Compute quantiles of distributions across each category of the cyclic granularity -->
<!-- 2. Compute JS distance between households for each each category of the cyclic granularity -->
<!-- 3. Total distance between households computed as sum of JS distances for all hours -->
<!-- 4. Cluster using this distance with hierarchical clustering algorithm (method "Ward.D") -->

@@

<!-- # ```{r range-parameter, fig.cap="The range of parameters used for the validation study, for the three different scenarios, number of simulations for each design, differences between means across granularities and series lengths. "} -->
<!-- # knitr::include_graphics("figs/simulation_table.png") -->
<!-- # ``` -->

@@

Huge data sets presents more complications for clustering. Clustering algorithms work well when there well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some households may have a mix of energy use patterns, which could be considered nuisance observations, between major clusters. For this reason, we have chosen to select a small group of households with relatively distinct and different patterns in order to illustrate the clustering, simply. This is often a good approach to tackling a big analysis task, so start with a simpler task. The approach, however, is applicable using on all households, but it would be more complicated to make final choice of the number of clusters and explain the results.

@@

<!-- # ```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in moy behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most households, there is one household for whom weekend behavior is standing out from the rest."} -->
<!-- # ``` -->

>>>>

\noindent This work is motivated by a need to cluster a large collection of residential smart meter data, so that customers can be grouped into similar energy usage patterns. These can be considered to be univariate time series of continuous values which are available at fine temporal scales. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and sporadic missing values. Using probability distributions is a natural way to analyze this types of data because they are robust to uneven length, missing data, or noise. This paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across a selection of cyclic granularities. Cyclic temporal granularities [@Gupta2021-hakear], which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can measure repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across cyclic granularities. The benefits of this approach are as follows.

@@

The primary application of this work is data from the Smart Grid, Smart City (SGSC) project (2010–2014) available through the [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). Half-hourly measurements of usage for more than 13,000 household electricity smart meters is provided from from October 2011 to March 2014. <!--Larger data sets include greater uncertainty about customer behavior due to growing variety of customers.--> Customers vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some customers use a dryer, while others dry their clothes on a line. Their weekly usage profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Daily energy usage varies depending on whether customers stay home or work away from home. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior complex. The challenge is to be able to cluster consumers into these type of expected patterns, and other unexpected patterns, using only their energy usage history [@Ushakova2020-rl]. <!--To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner.--> There is a growing need to have methods that can examine the energy usage heterogeneity observed in smart meter data and what are some of the most common power consumption patterns.

@@

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of customers showing similar periodic behavior by considering $wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Cluster characterization is an important final stage of a cluster analysis. The primary purpose is to compare the homogeneity within a cluster to the heterogeneity of clusters. This can be done numerically, by tabulating cluster means and standard deviations [@dasu2005grouping], and visually using methods for graphics multivariate data. @Cook2007-qe provides visual examples using both tours [@asimov1985grand] and parallel coordinate plots [@wegman1990hyperdimensional]. Dimension reduction techniques like principal component analysis [@jolliffe2016principal], multidimensional scaling (MDS) [@borg2005modern], t-distributed stochastic neighbor embedding (t-SNE) [@van2008visualizing] and linear discrimant analysis (LDA) [@fisher1936use] are also useful. <!-- A Parallel Coordinates Plot features parallel axes for each variable and each axis is linked by lines. Changing the axes may reveal patterns or relationships between variables for categorical variables. However, for categories with cyclic temporal granularities, preserving the underlying ordering of time is more desirable. Displaying cluster statistics is useful when we have larger problems and it is difficult to read the parallel coordinate plots due to congestion. All of MDS, PCA and t-SNE use a distance or dissimilarity matrix to construct a reduced-dimension space representation, but their goals are diverse. Multidimensional scaling [@borg2005modern) seeks to maintain the distances between pairs of data points, with an emphasis on pairings of distant points in the original space. The t-SNE embedding will compress data points that are close in high-dimensional space. Tour is a collection of interpolated linear projections of multivariate data into lower-dimensional space. The cluster characterization approach varies depending on the distance metric used. Parallel coordinate plots, scatter plot matrices, MDS or PCA are potentially useful ways to characterize clusters using wpd-based distances. For JS-based distances, plotting cluster statistics is beneficial for characterization and variable importance could be displayed through parallel coordinate plots. -->

@@

<!-- 1. Compute quantiles of distributions across each category of the cyclic granularity -->
<!-- 2. Compute JS distance between customers for each each category of the cyclic granularity -->
<!-- 3. Total distance between customers computed as sum of JS distances for all hours -->
<!-- 4. Cluster using this distance with hierarchical clustering algorithm (method "Ward.D") -->

@@

<!-- ```{r range-parameter, fig.cap="The range of parameters used for the validation study, for the three different scenarios, number of simulations for each design, differences between means across granularities and series lengths. "} -->
<!-- knitr::include_graphics("figs/simulation_table.png") -->
<!-- ``` -->

@@

Huge data sets presents more complications for clustering. Clustering algorithms work well when there well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some customers may have a mix of energy use patterns, which could be considered nuisance observations, between major clusters. For this reason, we have chosen to select a small group of customers with relatively distinct and different patterns in order to illustrate the clustering, simply. This is often a good approach to tackling a big analysis task, so start with a simpler task. The approach, however, is applicable using on all customers, but it would be more complicated to make final choice of the number of clusters and explain the results.

@@

<!-- # ```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in moy behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most customers, there is one household for whom weekend behavior is standing out from the rest."} -->
<!-- # ``` -->
